
## Case 8

Similarly, visualise the results for case 8. 

```{r}
load("data/RecurrenceCohort/cycIF/Final_cse8_data.RData")



immuneMarkers=c(2,  3 , 8 ,10, 11 ,13, 19 ,24, 28, 30, 34, 40)
luminalMarkers=c(4:7, 9, 12:14, 16:20,22, 25, 29,32:33, 36)

immlabels=c("bcell", "CD4", "cd8", "Tregs","macrophage", "pd1+")
stromaLabels=c("endothelial", "stroma", "vim+", "a-sma+")
lumLab2=c("CK+Ecad+" ,  "CK8 lo"  ,   "CKlow"    ,  "col-hi"   ,  "ER"   ,      "Ki67+"  ,    "myeop" ,     "basalLike")


case8_cellids$final2[which(case8_cellids$final2=="vim/sma")]="basalLike"
```

### Overview of the image

Plot showing the spatial distributions: We first focus on the epithelial cell types, then the immune cell types

```{r, fig.height=8, cache=T}
par(mfrow=c(1,2))

P1=brewer.pal(12, "Set3")
palette(P1)
#memVals=case8_cellids$final

memVals=case8_cellids$final2

searchType=lumLab2

## example whole slide 8DCIS
## DCIS
lx1=which(case8_cellids$type=="DCIS" & memVals%in%searchType)
lx2=which(case8_cellids$type=="DCIS" )
plot(case8_cellids$XLocN[lx2],max(case8_cellids$YLocN)-case8_cellids$YLocN[lx2], col="grey90", pch='.',  xlab="", ylab = "", main="8 DCIS peithelial") #xlim=c(12500, 14000), ylim=c(30500, 34000))
points(case8_cellids$XLocN[lx1],max(case8_cellids$YLocN)-case8_cellids$YLocN[lx1], col=factor(memVals[lx1], levels=searchType), main='DCIS', pch=20, xlim=c(15500, 22000), ylim=c(10000, 18000), cex=0.4)

## example whole slide 8IDC
## IDC

lx1=which(case8_cellids$type=="IDC" & memVals%in%searchType & case8_cellids$scene!="scene005")
lx2=which(case8_cellids$type=="IDC" & case8_cellids$scene!="scene005")
plot(case8_cellids$XLocN[lx2],max(case8_cellids$YLocN)-case8_cellids$YLocN[lx2], col="grey90", pch='.',  xlab="", ylab = "", main="8 DCIS immune") #xlim=c(12500, 14000), ylim=c(30500, 34000))
points(case8_cellids$XLocN[lx1],max(case8_cellids$YLocN)-case8_cellids$YLocN[lx1], col=factor(memVals[lx1], levels=searchType), main='DCIS', pch=20, xlim=c(15500, 22000), ylim=c(10000, 18000), cex=0.4)


#lx1=which(case8_cellids$type=="IDC" & memVals%in%searchType & case8_cellids$scene!="scene005")
#lx2=which(case8_cellids$type=="IDC" & case8_cellids$scene!="scene005")
#plot(case8_cellids$XLocN[lx2],max(case8_cellids$YLocN)-case8_cellids$YLocN[lx2], col="grey90", pch='.',  #xlab="", ylab = "",main="8 IDC peithelial"))
#points(case8_cellids$XLocN[lx1],max(case8_cellids$YLocN)-case8_cellids$YLocN[lx1], col=factor(memVals[lx1], #levels=searchType), main='DCIS', pch=20, cex=0.4)

## example whole slide 8DCIS immune
## DCIS
lx1=which(case8_cellids$type=="DCIS" & memVals%in%immlabels)
lx2=which(case8_cellids$type=="DCIS" )
plot(case8_cellids$XLocN[lx2],max(case8_cellids$YLocN)-case8_cellids$YLocN[lx2], col="grey90", pch='.',  xlab="", ylab = "", main="8 DCIS immune") #xlim=c(12500, 14000), ylim=c(30500, 34000))
points(case8_cellids$XLocN[lx1],max(case8_cellids$YLocN)-case8_cellids$YLocN[lx1], col=factor(memVals[lx1], levels=immlabels), main='DCIS', pch=20, xlim=c(15500, 22000), ylim=c(10000, 18000), cex=0.4)

## example whole slide 8IDC immune
## IDC
lx1=which(case8_cellids$type=="IDC" & memVals%in%immlabels & case8_cellids$scene!="scene005")
lx2=which(case8_cellids$type=="IDC" & case8_cellids$scene!="scene005")
plot(case8_cellids$XLocN[lx2],max(case8_cellids$YLocN)-case8_cellids$YLocN[lx2], col="grey90", pch='.',  xlab="", ylab = "", main="8 DCIS immune") #xlim=c(12500, 14000), ylim=c(30500, 34000))
points(case8_cellids$XLocN[lx1],max(case8_cellids$YLocN)-case8_cellids$YLocN[lx1], col=factor(memVals[lx1], levels=immlabels), main='DCIS', pch=20, xlim=c(15500, 22000), ylim=c(10000, 18000), cex=0.4)
```


### Compositional analysis


Quick overview of stroma, tumor and immune composition in dcis vs idc supports what we already know. DCIS has more immune cells compared to IDC

```{r}
case8_cellids$loc=paste(case8_cellids$type, case8_cellids$scene)
lx1=table(case8_cellids$loc[case8_cellids$final2%in%immlabels])
lx2=table(case8_cellids$loc[case8_cellids$final2%in%stromaLabels])
lx3=table(case8_cellids$loc[case8_cellids$final2%in%lumLab2])
LXAll=rbind(imm=lx1, str=lx2, ep=lx3)
Summ1=cbind(rowSums(LXAll[ ,1:10]), rowSums(LXAll[ ,11:14]))
Summ1=t(Summ1)/colSums(Summ1)
barplot(t(Summ1), col=c("#fdb462", "#984ea3","#8dd3c7"), names.arg=c("DCIS", "IDC"), horiz = T)
```

Barplots of specific cell types: CD8+ cells are a rare population but seems higher in DCIs compared to IDC. The proportion of Tregs increases in this transition.

Epithelial cells seems to have higher expression of CKs and Ecad in the DCIs compared to the IDC

```{r, fig.height=9}
par(mfrow=c(2,2))

lx1=table(case8_cellids$final2, case8_cellids$type)

lxLum=lx1[rownames(lx1)%in%lumLab2, ]
lxLumFrac=t(lxLum)/colSums(lxLum)
# 
 lxImm=lx1[rownames(lx1)%in%immlabels, ]
 lxImmFrac=t(lxImm)/colSums(lxImm)
 
 lxStr=lx1[rownames(lx1)%in%stromaLabels, ]
 lxStrFrac=t(lxStr)/colSums(lxStr)
 
 barplot(lxImmFrac, beside=T, ylab = "proportion of Immune cells", col=pal2sub, las=2)
 
 barplot(lxLumFrac[, lumLab2], beside=T, ylab = "proportion of Luminal cells", col=pal2sub, las=2)
 
 barplot(lxStrFrac, beside=T, col=pal2sub, ylab = "proportion of stromal cells", las=2)

```

Compare proportions between DCIS vs IDC? Here, need to omit the RBCs.
Perform a proport

```{r}
NcellsTotal=colSums(lx1[-match(c("omit", "rbc"), rownames(lx1)), ])
NImmuneTotal=colSums(lxImm)

sapply(1:nrow(lxImm), function(x) prop.test(lxImm[x, ], NImmuneTotal)$p.value)

# also test this for the tumo r component
NTumTotal=colSums(lxLum)

sapply(1:nrow(lxLum), function(x) prop.test(lxLum[x, ], NTumTotal)$p.value)

```


### Expression heatmap

Make a heatmap for the different clusters based on luminal vs stromal markers. The immune cell types are indicates in darker colors and match what was shown above

```{r, fig.height=8}
xa=c(lumLab2, immlabels)

CellList=sapply(xa, function(x) sample(which(case8_cellids$final2==x), 100))
#CellList=sapply(immlabels, function(x) sample(which(case8_cellids$final==x), 50))

cellAll=as.vector(CellList)
rowsideC=rep(c(pal1[1:length(lumLab2)], cols_d4[1:length(immlabels)]), each=100)

heatmap.2(case8_bck_mod[cellAll  ,luminalMarkers], col=RdBu[11:1],  scale="col", trace="none", RowSideColors = rowsideC, Rowv = NA)


heatmap.2(case8_bck_mod[cellAll  ,immuneMarkers], col=RdBu[11:1],  scale="col", trace="none", RowSideColors = rowsideC, Rowv = NA)

```

### Spatial statistics:

Due to the size of these images, these results were pre-run. Similar to case 9, we will plot:

* the interacting fraction
* the nearest neighbour distances
* the M-H interacting distances

#### The interacting fraction:

```{r}
# estblish the ppOut object

refineArea=which(case8_cellids$type=="IDC" & case8_cellids$scene!="scene005")
IFidc=InteractingFraction(memVals, refineArea, case8_cellids, labs=c("CD4", "cd8", "Tregs", "macrophage"), N=30)

xlimVd=c(min(case8_cellids$XLocN[refineArea]), max(case8_cellids$XLocN[refineArea]))
ylimVd=c(min(case8_cellids$YLocN[refineArea]), max(case8_cellids$YLocN[refineArea]))


refineArea=which(case8_cellids$type=="DCIS")
IFdcis=InteractingFraction(memVals, refineArea, case8_cellids, labs=c("CD4", "cd8", "Tregs", "macrophage"), N=30)

xlimVi=c(min(case8_cellids$XLocN[refineArea]), max(case8_cellids$XLocN[refineArea]))
ylimVi=c(min(case8_cellids$YLocN[refineArea]), max(case8_cellids$YLocN[refineArea]))


par(mfrow=c(1,2))
barplot(IFdcis$Fracfound, ylab="Percentage ''interacting'' with luminal", col=pal1[c(4, 3,2, 5)], main="8DCIS")
barplot(IFidc$Fracfound, ylab="Percentage ''interacting'' with luminal", col=palcols, main="8IDC")

```

We can compute a z-score by permuting the immune cell labels in order to determine whether a cell type is closer or further for a given composition. 

```{r}
par(mfrow=c(1,2))
load("data/RecurrenceCohort/cycIF/8dcis_spatial.Rdata")
dcisknn=knnDistMatrix
dcisdfall=cbind(dfall3, samp="dcis")
PlotInteractingZscore(IFdcis$Fracfound, DistMat, "8DCIS", col=pal1[c(4, 2:3, 5)])
load("data/RecurrenceCohort/cycIF/8idc_spatial.Rdata")
PlotInteractingZscore(IFidc$Fracfound, DistMat, "8IDC")
idcknn=knnDistMatrix
idcdfall=cbind(dfall3, samp="idc")
```

#### knn distance analysis

If we extend this distance, we can observe what the CDF of immune to epithelial interaction looks like: Firstly, for the 8DCIS cases. We use a boundary of 250 pixels from the border of the image to take into account cells which are too close to the border (and thus will artifically lengthen the interacting distance).

Firstly, we will look at the DCIS cases:

```{r}
Ctypes=levels(IFdcis$ppOut$marks)[-1]
ppOutR=IFdcis$ppOut

gridsize=250
ppInd=which(ppOutR$x>(xlimVd[1]+gridsize) & ppOutR$x<(xlimVd[2]-gridsize) & ppOutR$y>(ylimVd[1]+gridsize) & 
              ppOutR$y<(ylimVd[2]-gridsize))


par(mfrow=c(1, 3))
for (i in 1:3){
  i=1
  ptest=matrix(NA, nrow=4, ncol=4)
  namtest=matrix(NA, nrow=4, ncol=4)
  for (j in 1:(length(Ctypes)-1)){
    for (k in (j+1):length(Ctypes)){
    ptest[j,k]=ks.test(dcisknn[ which(ppOutR$marks[ppInd]==Ctypes[j]),1, i],
               dcisknn[ which(ppOutR$marks[ppInd]==Ctypes[k]),1, i])$p.value
      namtest[j,k]=paste(Ctypes[j], Ctypes[k])
    }}

  pvals=ptest[upper.tri(ptest)]
  names(pvals)=namtest[upper.tri(ptest)]

plot(ecdf(dcisknn[ which(ppOutR$marks[ppInd]=="Treg"),1, i]), lwd=2, main=sprintf("k=%s", i*2-1 ),
     xlim=c(0,300), col=palcols[1])
abline(v=30)
lines(ecdf(dcisknn[ which(ppOutR$marks[ppInd]=="CD8"),1, i]), lwd=2, col=palcols[2])
lines(ecdf(dcisknn[ which(ppOutR$marks[ppInd]=="CD4"),1, i]), lwd=2, col=palcols[3])
lines(ecdf(dcisknn[ which(ppOutR$marks[ppInd]=="Mac"),1, i]), lwd=2, col=palcols[4])
legend("bottomright", legend=paste(names(pvals), round(pvals*1000)/1000), pch=1)
}
```

Similarly, for 8 IDC:

```{r}
Ctypes=levels(IFidc$ppOut$marks)[-1]
ppOutR=IFidc$ppOut

gridsize=250
ppInd=which(ppOutR$x>(xlimVi[1]+gridsize) & ppOutR$x<(xlimVi[2]-gridsize) & ppOutR$y>(ylimVi[1]+gridsize) & 
              ppOutR$y<(ylimVi[2]-gridsize))

par(mfrow=c(1, 3))
for (i in 1:3){
  i=1
  ptest=matrix(NA, nrow=4, ncol=4)
  namtest=matrix(NA, nrow=4, ncol=4)
  for (j in 1:(length(Ctypes)-1)){
    for (k in (j+1):length(Ctypes)){
    ptest[j,k]=ks.test(idcknn[ which(ppOutR$marks[ppInd]==Ctypes[j]),1, i],
               idcknn[ which(ppOutR$marks[ppInd]==Ctypes[k]),1, i])$p.value
      namtest[j,k]=paste(Ctypes[j], Ctypes[k])
    }}

  pvals=ptest[upper.tri(ptest)]
  names(pvals)=namtest[upper.tri(ptest)]

plot(ecdf(idcknn[ which(ppOutR$marks[ppInd]=="Treg"),1, i]), lwd=2, main=sprintf("k=%s", i*2-1 ),
     xlim=c(0,300), col=palcols[1])
abline(v=30)
lines(ecdf(idcknn[ which(ppOutR$marks[ppInd]=="CD8"),1, i]), lwd=2, col=palcols[2])
lines(ecdf(idcknn[ which(ppOutR$marks[ppInd]=="CD4"),1, i]), lwd=2, col=palcols[3])
lines(ecdf(idcknn[ which(ppOutR$marks[ppInd]=="Mac"),1, i]), lwd=2, col=palcols[4])
legend("bottomright", legend=paste(names(pvals), round(pvals*1000)/1000), pch=1)
}

```

#### Morisita-Horn index of interaction

Below we combine the results of Morisita-Horn analysis for the different cell types of interest in the dcis and IDC samples. 
Here, we look at the values if we 

* shift the grid
* omit both squares containing only immune or only tumor squares (ie. focusing on the immune-tumor boundary)
* omit squares containing only the immune cells
* omit squares containing only tumor cells

We use the last example in publication


```{r}
all8mh=rbind(dcisdfall, idcdfall)
#all8mh=all8mh[which(all8mh$type=="shift_omit_tum"), ]

ggplot(all8mh, aes(x=comp, y=mean, col=samp))+facet_grid(~type)+geom_point()+geom_errorbar(aes(ymin=lower, ymax=upper))+theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
all8sam2=all9sam[which(all8mh$type=="shift_omit_tum"), ]

lx2=unique(all8sam2$comp)

Pval=rep(NA, length(unique(lx2)))

for (i in 1:length(lx2)){
  ax1=which(all8sam2$comp==lx2[i])
  sde=(all8sam2$upper[ax1[2]]-all8sam2$lower[ax1[2]])/(2*1.96)
  zv=abs(all8sam2$mean[ax1[2]]-all8sam2$mean[ax1[1]])/sde
  Pval[i]=exp(-0.717*zv -0.416*zv^2)
}

names(Pval)=unique(lx2)
p.adjust(Pval)
```

## Summary of the composition of case8 and case 9

(for use with frequencies in cibersort)

```{r}
lxImmFracTot=data.frame(rbind(lxImmFrac9[ ,c(1:5)], lxImmFrac[ ,c(1:3, 6,5)]))
lxImmFracTot$Type=rep(c("DCIS", "IDC"), 2)
lxImmFracTot$Case=rep(c("Case9", "Case8"),each=2)
      
lxImmFracTotmelt=melt(lxImmFracTot, id.vars = c("Type", "Case"))

ggplot(lxImmFracTotmelt, aes(x=Type, y=value, col=Case))+geom_point()+geom_line(aes(group=Case))+facet_wrap(~variable, scale="free_y")+
  scale_color_manual(values=c("#619CFF", "#A58AFF"))

```