# Analysis of H \& E images

Below is the code used to analyse the H\&E tissue sections. As an example, code for this analysis is shown for case 9, and the same code is exmployed for all other samples

## Macrodissecting regions

In each tissue slide, view this within R and macrodissect the points within this which are confined within the DCIS, IDC and normal regions: the user is prompted to draw a box around these, and the area of interest is saved in a ROIlist. These are saved to file and are used in subsequent analysis. 

6 regions of interest are determined in case 9IDC:
1. Is a IDC region
2. IDC
3. DCIS like
4. DCIS like
5. normal
6. normal

Cells are color coded: immune is orange, tumor cells are green and stromal cells are purple.

```{r, eval=F}
idx=12
savNames="9IDC"
NROI=6 ## number of regions of interest intended (exclude entire slide)
roiNames=c("idc1","idc2", "dcis1","dcis2", "n1","n2", "all")

X11()
plot(QPs[[idx]]$CentroidX, QPs[[idx]]$CentroidY, col=QPs[[idx]]$CellType, pch='.')

## check if this works, otherwise may manually need to input this data
ROIlist=list()
for (i in 1:NROI){
ROIlist[[i]]<- fhs(QPs[[idx]][ ,3:4])
}

names(ROIlist)=roiNames[1:NROI]

save(ROIlist,Ntab, file=paste(savNames, "extracted_points_qupath.RData", sep="_"))
```

```{r}
palette(c("#fdb462", "#984ea3","#8dd3c7"))

load("data/RecurrenceCohort/qupath/9IDC_extracted_points_qupath.RData")
idx=12
savNames="9IDC"

gridsize=100
NROI=6 ## number of regions of interest intended (exclude entire slide)
roiNames=c("idc1","idc2", "dcis1","dcis2", "n1","n2", "all")


xlimV=c(min(QPs[[idx]]$CentroidX), max(QPs[[idx]]$CentroidX))
ylimV=c(min(QPs[[idx]]$CentroidY), max(QPs[[idx]]$CentroidY))

xsize=diff(xlimV)/1000
ysize=diff(ylimV)/1000

xcut=seq(xlimV[1], xlimV[2], length=round(diff(xlimV)/gridsize)+1)
ycut=seq(ylimV[1], ylimV[2], length=round(diff(ylimV)/gridsize)+1)

windowOut=ripras(QPs[[idx]]$CentroidX, QPs[[idx]]$CentroidY)
ppOutR=ppp(QPs[[idx]]$CentroidX, QPs[[idx]]$CentroidY, marks=QPs[[idx]]$CellType, poly=windowOut$bdry)

ppTess=tess(ppOut, xgrid=xcut, ygrid=ycut)

nt=quadratcount(ppOutR[ppOutR$marks=="Tumor"], tess=ppTess)
ni=quadratcount(ppOutR[ppOutR$marks=="Immune cells"], tess=ppTess)
ns=quadratcount(ppOutR[ppOutR$marks=="Stroma"], tess=ppTess)
na=quadratcount(ppOutR, tess=ppTess)

bd=list()
for (i in 1:NROI){
  bd[[i]]=ripras(QPs[[idx]]$CentroidX[as.numeric(ROIlist[[i]])] , QPs[[idx]]$CentroidY[as.numeric(ROIlist[[i]])])
}

plot(QPs[[idx]]$CentroidX, QPs[[idx]]$CentroidY, col=QPs[[idx]]$CellType, pch='.', xlim=c(2000, 6000), ylim=c(17000, 20000))
for (i in 1:length(bd)){
  polygon(bd[[i]]$bdry[[1]]$x, bd[[i]]$bdry[[1]]$y)
 # text(mean(bd[[i]]$xrange), mean(bd[[i]]$yrange), i)
}

# Create a mask for later analysis

Mask=array(numeric(), c(nrow(nt), ncol(nt), NROI))

for (i in 1:NROI){
  xidc1=unique(tileindex(QPs[[idx]]$CentroidX[as.numeric(ROIlist[[i]])], QPs[[idx]]$CentroidY[as.numeric(ROIlist[[i]])], ppTess))
  xidc1b=strsplit(as.character(xidc1), ", col ")
  xvals=as.numeric(sapply(xidc1b, function(x) substr(x[1], 10, 13)))
  yvals=as.numeric(sapply(xidc1b, function(x) x[2]))
  tempMat=matrix(0, ncol=ncol(Mask), nrow=nrow(Mask))
    for (j in 1:length(xvals)){
    tempMat[xvals[j], yvals[j]]=1
    }
   Mask[, , i]=tempMat
}
```

## Compositional summary

The proportion of cells within each ROI is shown below for Case 9:

```{r}

Ntab=sapply(ROIlist, function(x) table(QPs[[idx]][x ,2]))
Ntab=cbind(Ntab, all=table(QPs[[idx]][ ,2]))

CellSummN=t(Ntab)/colSums(Ntab)
CellSummN2=CellSummN[NROI+1, ]

barplot(t(CellSummN), col=c("#fdb462", "#984ea3","#8dd3c7"), main="Case 9 proportion of cells at each site")
```

Accross all the patient samples, we can assess what the composition is like in each H\&E section. The regions are annotated as:

* dd: dcis region within a H\&E section diagnosed as DCIS
* id: adjacent dcis within a diagnosed IDC
* ii: idc region within a diagnosed IDC.

Below are the raw number of cells in each region

```{r}
SummData=read_xlsx("data/RecurrenceCohort/qupath/cellcounts_HE.xlsx", sheet=2)
lx2=which(SummData$Sample%in%c("dcis", "idc", "dcisAll", "idcAll"))
SummData2=SummData[lx2, c(1:5)]
SummData2melt=melt(SummData2)
SummData2melt$Sample2=paste( substr(SummData2melt$Case, 2, 2), substr(SummData2melt$Sample, 1, 1))
SummData2melt$Case2=substr(SummData2melt$Case,1,1)
palette(c("#fdb462", "#984ea3","#8dd3c7", "#e41a1c", "#377eb8", "#f781bf", "#999999"))

ggplot(SummData2melt, aes(x=Sample2, y=value, fill=variable))+geom_bar(stat="identity")+facet_grid(~Case2)+scale_fill_manual(values=c("#fdb462", "#984ea3","#8dd3c7", "#e41a1c", "#377eb8", "#f781bf", "#999999"))
```

For this study, we focus only on DCIS regions within a clincailly diagnosed speciment DCIS, and IDC regions within a clinically diagnosed IDC specimen. The proportions of these different cells are shown below:

```{r, fig.height=6}
rmX=grep("dcis", SummData2$Sample)
rmY=grep("idc", SummData2$Case)
rmZ=intersect(rmX, rmY)

SD2=SummData2[-rmZ, ]
SD3=SD2[ ,3:5]/rowSums(SD2[ ,3:5])

colnames(SD3)=colnames(SD2)[3:5]
rownames(SD3)=t(SD2[ ,2])
barplot(t(SD3), col=c("#fdb462", "#984ea3","#8dd3c7"))

```

Combine the over data in one plot and computer differences in proportion

```{r}
SD3$sample=rownames(SD3)
SD3$tumor=substr(rownames(SD3), 2, 4)
SD3$case=substr(rownames(SD3), 1, 1)
sd4=melt(SD3)
ggplot(sd4, aes(x=tumor, y=value, col=case))+geom_point()+geom_line(aes(group=case))+facet_grid(~variable)

# for each individual pair
Ptets=sapply(1:6, function(x) prop.test(t(SD2[(x*2-1):(x*2), 3 ]), rowSums(SD2[(x*2-1):(x*2), 3:5 ])))

# # proportion test
# AllSamples=prop.test(t(SD2[, 3 ]), rowSums(SD2[, 3:5 ]))

# by beta regression?
p_test=betareg(Immune~tumor+case, data=SD3[SD3$case!=1, ])
p_test=betareg(Tumor~tumor+case, data=SD3[SD3$case!=1, ])
p_test=betareg(Stroma~tumor+case, data=SD3[SD3$case!=1, ])
# paired t test
t.test(SD3$Immune[SD3$tumor=="dci"][1:5],SD3$Immune[SD3$tumor=="idc"][1:5], paired=T, alternative="greater")
t.test(SD3$Tumor[SD3$tumor=="dci"][1:5],SD3$Tumor[SD3$tumor=="idc"][1:5], paired=T, alternative="greater")
```

## Spatial statistics: based on nearest neighbour distances

Using these points, we can conduct spatial statistics using the points in these regions using the spatsats package.

```{r, eval=F}
xlimV=c(min(QPs[[idx]]$CentroidX), max(QPs[[idx]]$CentroidX))
ylimV=c(min(QPs[[idx]]$CentroidY), max(QPs[[idx]]$CentroidY))

xsize=diff(xlimV)/1000
ysize=diff(ylimV)/1000

xcut=seq(xlimV[1], xlimV[2], length=round(diff(xlimV)/gridsize)+1)
ycut=seq(ylimV[1], ylimV[2], length=round(diff(ylimV)/gridsize)+1)

windowOut=ripras(QPs[[idx]]$CentroidX, QPs[[idx]]$CentroidY)
ppOutR=ppp(QPs[[idx]]$CentroidX, QPs[[idx]]$CentroidY, marks=QPs[[idx]]$CellType, poly=windowOut$bdry)

ppTess=tess(ppOut, xgrid=xcut, ygrid=ycut)

nt=quadratcount(ppOutR[ppOutR$marks=="Tumor"], tess=ppTess)
ni=quadratcount(ppOutR[ppOutR$marks=="Immune cells"], tess=ppTess)
ns=quadratcount(ppOutR[ppOutR$marks=="Stroma"], tess=ppTess)
na=quadratcount(ppOutR, tess=ppTess)

bd=list()
for (i in 1:NROI){
  bd[[i]]=ripras(QPs[[idx]]$CentroidX[as.numeric(ROIlist[[i]])] , QPs[[idx]]$CentroidY[as.numeric(ROIlist[[i]])])
}
```

### Nearest neighbor distances

One of the most straigh-forwards ways of lookig for spatial differences is by calculating nearest neighbour distances between each immune cell to it's nearest tumor cell. We can also explore how much this metric varies wehen we consider 1 nearest neighbour, 3 nearest neighbours or 5 nearest neighbours.

Below are the CDFs of the distances for the above segmented regions, light pink are the dcis regions, maroon are idc and blue are the normal regions.

Example code for calculating nearest neighbor distances in case 9. We can explore differences between k=1, 3 and 5. Differences between two curves is calculated by the KS statistic.

Both dcis and idc appear to have shorter immune-tumor/epithelial distances compared to the normals.

```{r, cache=T}
amx1=c("Immune", "Stroma", "Tumor")

lx1=grep("dcis", roiNames) 
cx1=brewer.pal(9, "RdPu")[ 2:(length(lx1)+1)]
lx2=grep("idc", roiNames)
cx2=brewer.pal(9, "PuRd")[ 9:(10-length(lx2))]
lx3=grep("n", roiNames)
cx3=brewer.pal(9, "BuPu")[2:(length(lx1)+1)]
palCols=rep(NA, length(roiNames))
palCols[lx1]=cx1
palCols[lx3]=cx3
palCols[lx2]=cx2

palette(palCols)

d2=nndist(ppOutR, by=marks(ppOutR), k=c(1:5))
Dist1=d2[ ,grep("dist.1", colnames(d2))]
knn3=sapply(amx1, function(x) rowMeans(d2[ ,grep(x, colnames(d2))[1:3]]))
knn5=sapply(amx1, function(x) rowMeans(d2[ ,grep(x, colnames(d2))]))
knnDistMatrix=abind(Dist1, knn3, knn5, along=3)

par(mfrow=c(1,3))
for (i in 1:3){
  if (NROI>1){
# compare the different lines?
  ptest=matrix(NA, nrow=NROI, ncol=NROI)
  namtest=matrix(NA, nrow=NROI, ncol=NROI)
  for (j in 1:(NROI-1)){
    for (k in (j+1):NROI){
      ptest[j,k]=ks.test(knnDistMatrix[ which(ppOutR$marks[as.numeric(ROIlist[[j]])]=="Immune cells"),3, i],
                         knnDistMatrix[ which(ppOutR$marks[as.numeric(ROIlist[[k]])]=="Immune cells"),3, i])$p.value
      namtest[j,k]=paste(roiNames[j], roiNames[k])
    }}

  pvals=ptest[upper.tri(ptest)]
  names(pvals)=namtest[upper.tri(ptest)]
  }

plot(NA, xlim=c(0, 200), ylim=c(0, 1), main=sprintf("Immune-Tum k=%s", i*2-1 ), ylab="CDF", xlab="Nearest stromal distance")
## plot the "stromal distances"?
for (j in 1:NROI){
  lines(ecdf(knnDistMatrix[ which(ppOutR$marks[as.numeric(ROIlist[[j]])]=="Immune cells"),3, i]), lwd=1, col=j)
}
legend("topright", legend=roiNames, col=c(1:NROI), pch=19)

if (NROI>1){
legend("bottomright", legend=paste(names(pvals), round(pvals*1000)/1000), pch=1)
}
}

```

We can also perform the same analysis for the immune-stroma fraction. In all these cases there is virtually no difference between 

```{r, cache=T}
par(mfrow=c(1,3))
for (i in 1:3){
if (NROI>1){
ptest=matrix(NA, nrow=NROI, ncol=NROI)
namtest=matrix(NA, nrow=NROI, ncol=NROI)
for (j in 1:(NROI-1)){
  for (k in (j+1):NROI){
    ptest[j,k]=ks.test(knnDistMatrix[ which(ppOutR$marks[as.numeric(ROIlist[[j]])]=="Immune cells"),2, i],
                       knnDistMatrix[ which(ppOutR$marks[as.numeric(ROIlist[[k]])]=="Immune cells"),2, i])$p.value
    namtest[j,k]=paste(roiNames[j], roiNames[k])
  }}

pvals=ptest[upper.tri(ptest)]
names(pvals)=namtest[upper.tri(ptest)]
}

plot(NA, xlim=c(0, 200), ylim=c(0, 1), main=sprintf("Immune-Stroma k=%s", i*2-1 ), ylab="CDF", xlab="Nearest stromal distance")
## plot the "stromal distances"?
for (j in 1:NROI){
  lines(ecdf(knnDistMatrix[ which(ppOutR$marks[as.numeric(ROIlist[[j]])]=="Immune cells"),2, i]), lwd=1, col=j)
}

if (NROI>1){
legend("bottomright", legend=paste(names(pvals), round(pvals*1000)/1000), pch=1)
}
legend("topright", legend=roiNames, col=c(1:NROI), pch=19)
}
```


Compare nearest neighbour distances for the different cases: 

* immune - immune
* immune - stroma
* immune - tumor

```{r}
load("data/RecurrenceCohort/qupath/auc_knn_region_segmented.RData")
Inputs1=sapply(OutputKnnAUC, function(x) x[ ,1])
Inputs2=sapply(OutputKnnAUC, function(x) x[ ,2])
Inputs3=sapply(OutputKnnAUC, function(x) x[ ,3])

par(mfrow=c(1,3))
boxplot(Inputs2[1, -c(1:7)]~substr(colnames(Inputs1)[-c(1:7)],2,3), main="immune", col=dcisidcCol, ylab="AUC scores 50um")
boxplot(Inputs2[2,-c(1:7)]~substr(colnames(Inputs1)[-c(1:7)],2,3), main="stroma",col=dcisidcCol)
boxplot(Inputs2[3,-c(1:7)]~substr(colnames(Inputs1)[-c(1:7)],2,3), main="tumor", col=dcisidcCol)

# wilcox test here
wilcox.test(Inputs2[3, -c(1:7)]~substr(colnames(Inputs1)[-c(1:7)],2,3), main="immune",  ylab="AUC scores 50um")
wilcox.test(Inputs2[1, -c(1:7)]~substr(colnames(Inputs1)[-c(1:7)],2,3), main="immune",  ylab="AUC scores 50um")
wilcox.test(Inputs2[2, -c(1:7)]~substr(colnames(Inputs1)[-c(1:7)],2,3), main="immune",  ylab="AUC scores 50um")

# can also report beta-regression score
summary(betareg(Inputs2[3, -c(1:7)]~substr(colnames(Inputs1)[-c(1:7)],2,3)))#, main="immune",  ylab="AUC scores 50um")
summary(betareg(Inputs2[2, -c(1:7)]~substr(colnames(Inputs1)[-c(1:7)],2,3)))
summary(betareg(Inputs2[1, -c(1:7)]~substr(colnames(Inputs1)[-c(1:7)],2,3)))
```


### The interacting fraction

We can define the interacting fraction by drawing a vertical line at a given distance R, and finding the intersection with the y-axis in the CDF plot. I.e. this is the porportion of immune cells which are within a distance R of a tumor cell. 

We can collate the values for all samples below:

```{r}
qpathdir="data/RecurrenceCohort/qupath/"
mx1=dir(qpathdir, "knn_mh_outputs*")

snames=sapply(strsplit(mx1, "_"), function(x) x[4])


Allknn=list()
AllMH=list()

AllROI=list()
DistZscores=list()

AllTIdist=list()
AllITdist=list()

AllTInorm=list()
AllITnorm=list()

ncountD=1
repidx=NA

for (i in 1:length(mx1)){
  load(paste(qpathdir, mx1[i], sep=""))
  a1=grep("DCIS", snames[i])
  a2=grep("IDC", snames[i])
  
  AllROI[[i]]=ROIlist
  DistZscores[[i]]=Zscores
  AllMH[[i]]=dfall3
  Allknn[[i]]=knnDistMatrix
  
  if (length(a1)==1){
    x1=grep("dcis", names(ITdist))
  #  AllTIdist[[i]]=unlist(TIdist[x1])
    AllITdist[[i]]=unlist(ITdist[x1])
  
  }
   if (length(a2)==1){
    x1=grep("idc", names(ITdist))
    AllITdist[[i]]=unlist(ITdist[x1])
  #  AllTIdist[i]=unlist(TIdist[x1])
   
   }
   a3=grep("n", names(ITdist))
 #  AllTInorm[[i]]=unlist(TIdist[a3])
   AllITnorm[[i]]=unlist(ITdist[a3])
}


names(AllITdist)=paste(snames, 1)#paste(rep(snames, times=repidx), unlist(sapply(repidx, function(x) seq(1, x))))
names(AllITnorm)=paste(snames, 1)
```


```{r}
intDistance=10
PropIT=sapply(AllITdist, function(x) length(which(x<intDistance)))/sapply(AllITdist, length)
#PropTI=sapply(AllTIdist, function(x) length(which(x<intDistance)))/sapply(AllTIdist, length)

AllIT=sapply(AllITnorm, function(x) length(which(x<intDistance)))/sapply(AllITnorm, length)
#AllTI=sapply(AllTInorm, function(x) length(which(x<intDistance)))/sapply(AllTInorm, length)

Prop2=rbind(PropIT, AllIT)
#colnames(Prop2)=snames

Prop2Melt=melt(Prop2)
Prop2Melt$Pat=substr(Prop2Melt$Var2, 1, 1)
Prop2Melt$Reg=substr(Prop2Melt$Var2, 2, 2)

ggplot(Prop2Melt, aes(x=Reg, y=value, fill=Var1))+geom_bar(stat="identity", position="dodge")+facet_grid(~Pat)+ggtitle("Proportion of cells within interacting distance")+scale_fill_discrete(name = "Sample", labels = c("tumor (dcis idc)", "Normal"))+xlab("Region (DCIS or IDC HE slide")+ylab("Proportion")

ggplot(Prop2Melt[Prop2Melt$Var1=="PropIT", ], aes(x=Reg, y=value, col=Pat))+geom_point()+geom_line(aes(group=Pat))+ggtitle("Proportion of cells within interacting distance")+xlab("Region (DCIS or IDC HE slide")+ylab("Proportion")


test1=betareg(value~Reg+Pat, data=Prop2Melt[Prop2Melt$Var1=="PropIT", ])

t.test(Prop2Melt[Prop2Melt$Var1=="PropIT" & Prop2Melt$Pat!=1 , "value"] ~
         Prop2Melt[Prop2Melt$Var1=="PropIT"& Prop2Melt$Pat!=1, "Reg"], paired=T)
```


## Morisita-Horn metric

The Morisita-Horn index can be interpreted as a spatial correlation coefficient between two cell types within an image. Here, we consider the immune population and the tumor population.

For the whole image using boostrap method:

```{r, cache=T}
## need to set seed for 95% CI. There is too much variation in these values

allDat=data.frame(Tumor=as.vector(nt), Immune=as.vector(ni), Stroma=as.vector(ns))
xres <-mh(allDat, resample=500)

# refine this to regions of interest
mhRegion=function(nt, ni, ns, mask, resample=500){
  idc1Dat=data.frame(Tumor=as.vector(nt)*as.vector(mask), Immune=as.vector(ni)*as.vector(mask), Stroma=as.vector(ns)*as.vector(mask))
  rmidx=which(is.na(idc1Dat[ ,1])==T)
  
  if (length(rmidx>0)){
  idc1Dat=idc1Dat[-rmidx, ]
  }
  lx1=colSums(idc1Dat)
  lx2=which(lx1==0)
  if (length(lx2)>0){
    idc1Dat[ ,lx2]=1
  }
  xres2 <-mh(idc1Dat, resample=resample)
  if (length(lx2)>0){
    xres2$Mean[lx2, ]=NA
    xres2$Lower.Quantile[lx2, ]=NA
    xres2$Upper.Quantile[lx2, ]=NA
  }
  xres2
}

MHsumm=matrix(NA, nrow=3, ncol=3*(NROI+1))

for (i in 1:NROI){
  MH1=mhRegion(nt, ni, ns, Mask[ , , i])
  MHsumm[ 1,(3*i-2):(3*i)]=MH1$Mean[lower.tri(MH1$Mean)]
  MHsumm[ 2,(3*i-2):(3*i)]=MH1$Lower.Quantile[lower.tri(MH1$Lower.Quantile)]
  MHsumm[ 3,(3*i-2):(3*i)]=MH1$Upper.Quantile[lower.tri(MH1$Upper.Quantile)]
}

MHsumm[ 1,(3*(NROI+1)-2):(3*(NROI+1))]=xres$Mean[lower.tri(xres$Mean)]
MHsumm[ 2,(3*(NROI+1)-2):(3*(NROI+1))]=xres$Lower.Quantile[lower.tri(xres$Lower.Quantile)]
MHsumm[ 3,(3*(NROI+1)-2):(3*(NROI+1))]=xres$Upper.Quantile[lower.tri(xres$Upper.Quantile)]

Samp=rep(roiNames, each=3)
Comp=rep(c("TI", "ST", "SI"), (NROI+1))

dfall=data.frame(Samp=Samp, Comp=Comp, MH=MHsumm[ 1,], MHlow=MHsumm[2,], MHhi=MHsumm[3,])
```

Plot the output for case 9 for example:

```{r}
ggplot(dfall, aes(x=Samp, y=MH, col=Comp))+geom_point()+geom_errorbar(aes(ymin=MHlow, ymax=MHhi), width=.2)
```

We can also test the stability but gradually shifting the grid to see what the error margin is like using the below code (Note this is not run due to time constraints). In addition, we can remove grids which only have "tumor" cells, in order to evaluate the interaction in the stroma and at the tumor-stroma interface in order to account for some differences in tissue architecture between dcis and idc.

```{r, eval=F}
distx=seq(0, gridsize/2, by=2)
disty=seq(0,  gridsize/2, by=2)

distXb=seq( gridsize/2, 0, by=-2)
distYb=seq( gridsize/2, 0, by=-2)

MHsummGrid=matrix(NA, nrow=(length(distx)^2), ncol=3*NROI)
MHsummNoTum=matrix(NA, nrow=(length(distx)^2), ncol=3*NROI)
MHsummNoImmstr=matrix(NA, nrow=(length(distx)^2), ncol=3*NROI)
MHsummNoTumImm=matrix(NA, nrow=(length(distx)^2), ncol=3*NROI)
MHsummNoImm=matrix(NA, nrow=(length(distx)^2), ncol=3*NROI)

count=1

for (i in 1:length(distx)){ #length(distx)
  for (j in 1:length(distXb)){ #length(distXb)
    txlim=c(xlimV[1]+distx[i], xlimV[2]-distXb[i])
    tylim=c(ylimV[1]+disty[j], ylimV[2]-distYb[j]) 
    
    xcut=seq(txlim[1], txlim[2], length=round(diff(c(txlim))/gridsize)+1)
    ycut=seq(tylim[1], tylim[2], length=round(diff(c(tylim))/gridsize)+1)
    
    # cut into quadrants
    dfT=QPs[[idx]][which(QPs[[idx]]$CentroidX >txlim[1] & QPs[[idx]]$CentroidX <txlim[2] & 
                           QPs[[idx]]$CentroidY > tylim[1] & QPs[[idx]]$CentroidY < tylim[2] ), ]
    ppOut=ppp(dfT$CentroidX, dfT$CentroidY, marks=dfT$CellType, poly=windowOut$bdry)
    ppTessT=tess(ppOut, xgrid=xcut, ygrid=ycut)
    nt=quadratcount(ppOut[ppOut$marks=="Tumor"], tess=ppTessT)
    ni=quadratcount(ppOut[ppOut$marks=="Immune cells"], tess=ppTessT)
    ns=quadratcount(ppOut[ppOut$marks=="Stroma"], tess=ppTessT)
    
    # Create a Mask for each
    Mask=array(numeric(), c(nrow(nt), ncol(nt), NROI))
    for (k in 1:NROI){
      xidc1=unique(tileindex(QPs[[idx]]$CentroidX[as.numeric(ROIlist[[k]])], QPs[[idx]]$CentroidY[as.numeric(ROIlist[[k]])], ppTessT))
      xidc1b=strsplit(as.character(xidc1), ", col ")
      xvals=as.numeric(sapply(xidc1b, function(x) substr(x[1], 10, 13)))
      yvals=as.numeric(sapply(xidc1b, function(x) x[2]))
      tempMat=matrix(0, ncol=ncol(Mask), nrow=nrow(Mask))
      for (l in 1:length(xvals)){
        tempMat[xvals[l], yvals[l]]=1
      }
      Mask[, , k]=tempMat
    }
    
    # figure out allDat for each ROI
    allDat2=lapply(1:NROI, function(x) data.frame(Tumor=as.vector(nt)*as.vector(Mask[, , x]), 
                                                          Immune=as.vector(ni)*as.vector(Mask[, , x]), 
                                                          Stroma=as.vector(ns)*as.vector(Mask[, , x])))
    allDat3<-lapply(allDat2, function(x) x[-which(rowSums(x)==0), ])
    
    # use all points
    xres<-lapply(allDat3, function(x) mh(x[ ,1:2], resample=1))
    MHsummGrid[count, ]=as.vector(sapply(xres, function(x) x$Mean[lower.tri(x$Mean)]))
    allDat4<-lapply(allDat3, function(x) x[-which(rowSums(x[ ,2:3])<2), ])
    cidx=sapply(allDat4, nrow)
    ax=which(cidx==0)
    if (length(ax)>0){
      allDat4[ax]=allDat3[ax]
    }
    xres<-lapply(allDat4, function(x) mh(x[ ,1:2], resample=1))
    MHsummNoImmstr[count, ]=as.vector(sapply(xres, function(x) x$Mean[lower.tri(x$Mean)]))
    # rm Tum==0
    allDat4<-lapply(allDat3, function(x)  x[-which(x[ ,1]<2), ])
    cidx=sapply(allDat4, nrow)
    ax=which(cidx==0)
    if (length(ax)>0){
      allDat4[ax]=allDat3[ax]
    }
    xres<-lapply(allDat4, function(x) mh(x[ ,1:2], resample=1))
    MHsummNoTum[count, ]=as.vector(sapply(xres, function(x) x$Mean[lower.tri(x$Mean)]))
    # rm Imm = 0  and Tum =0
    allDat4<-lapply(allDat3, function(x) x[-which(rowSums(x[ ,1:2])<2), ])
    cidx=sapply(allDat4, nrow)
    ax=which(cidx==0)
    if (length(ax)>0){
      allDat4[ax]=allDat3[ax]
    }
    xres<-lapply(allDat4, function(x) mh(x[, 1:2], resample=1))
    MHsummNoTumImm[count, ]=as.vector(sapply(xres, function(x) x$Mean[lower.tri(x$Mean)]))
    
    # rm Imm = 0 only
    allDat4<-lapply(allDat3, function(x) x[-which(x[ ,2]<2), ])
    cidx=sapply(allDat4, nrow)
    ax=which(cidx==0)
    if (length(ax)>0){
      allDat4[ax]=allDat3[ax]
    }
    xres<-lapply(allDat4, function(x) mh(x[ ,1:2], resample=1))
    MHsummNoImm[count, ]=as.vector(sapply(xres, function(x) x$Mean[lower.tri(x$Mean)]))
    
    count=count+1
  }
}

colnames(MHsummGrid)=paste(rep(roiNames, each=3), rep(c("imm-tum", "str-tum", "str-imm"), NROI), sep="-")
colnames(MHsummNoTum)=colnames(MHsummGrid)
colnames(MHsummNoImmstr)=colnames(MHsummGrid)
colnames(MHsummNoTumImm)=colnames(MHsummGrid)
colnames(MHsummNoImm)=colnames(MHsummGrid)


MHnoim=sapply(1:ncol(MHsummNoImm), function(x) quantile(MHsummNoImm[ ,x], c(0.025, 0.975), na.rm = T))
dfallc=cbind(colMeans(MHsummNoImm, na.rm=T), MHnoim[1, ], MHnoim[2, ])
dfallc=data.frame(dfallc)
dfallc$comp=colnames(MHsummGrid)
colnames(dfallc)=c("mean", "lower", "upper", "comp")
dfallc$type="Rm imm = 0"

MHrall=sapply(1:ncol(MHsummGrid), function(x) quantile(MHsummGrid[ ,x], c(0.025, 0.975), na.rm = T))
dfall=cbind(colMeans(MHsummGrid, na.rm=T), MHrall[1, ], MHrall[2, ])
dfall=data.frame(dfall)
dfall$comp=colnames(MHsummGrid)
colnames(dfall)=c("mean", "lower", "upper", "comp")
dfall$type="keep all"


MHexcBoth=sapply(1:ncol(MHsummGrid), function(x) quantile(MHsummNoTumImm[ ,x], c(0.025, 0.975), na.rm = T))
dfallb=cbind(colMeans(MHsummNoTumImm, na.rm=T), MHexcBoth[1, ], MHexcBoth[2, ])
dfallb=data.frame(dfallb)
dfallb$comp=colnames(MHsummGrid)
colnames(dfallb)=c(  "mean", "lower", "upper", "comp")
dfallb$type="rm imm=0 or tum=0"

MHexcImm=sapply(1:ncol(MHsummGrid), function(x) quantile(MHsummNoImmstr[ ,x], c(0.025, 0.975), na.rm = T))
dfalli=cbind(colMeans(MHsummNoImmstr, na.rm=T), MHexcImm[1, ], MHexcImm[2, ])
dfalli=data.frame(dfalli)
dfalli$comp=colnames(MHsummGrid)
colnames(dfalli)=c(  "mean", "lower", "upper", "comp")
dfalli$type="rm_imm_str0"


MHexcTum=sapply(1:ncol(MHsummGrid), function(x) quantile(MHsummNoTum[ ,x], c(0.025, 0.975), na.rm = T))
dfallg=cbind(colMeans(MHsummNoTum, na.rm=T), MHexcTum[1, ], MHexcTum[2, ])
dfallg=data.frame(dfallg)
dfallg$comp=colnames(MHsummGrid)
colnames(dfallg)=c(  "mean", "lower", "upper", "comp")
dfallg$type="rm_tum=0"
dfall3=rbind(dfall, dfallg, dfalli, dfallb, dfallc)
```


Using this process, we can compute an average and estimate the error for this process. Below, we report the MH-index across the patient samples. Firstly, we can visualise the immune-tumor interactions. The normal samples seem to have a higher interaction between immune and epithelial cells. 


```{r}
# patha="~/Desktop/spatstat_rslt/"
# path1=dir(patha, "mh_analysis.RData")


names(AllMH)=snames
MHmelt=melt(AllMH, measure.vars="mean") 
MHmelt2=MHmelt[grep("imm-tum", MHmelt$comp), ]
MHmelt2$nam=substr(MHmelt2$L1, 1, 1)
MHmelt2$class=substr(MHmelt2$L1, 2, 2)
MHmelt2$sampType=substr(MHmelt2$comp, 1, 1)
MHmelt2=MHmelt2[-which(MHmelt2$class=="I" & MHmelt2$sampType=="d"), ]
#MHmelt2$sampType2=paste(MHmelt2$comp, MHmelt$type)
MHmelt3=MHmelt2[which(MHmelt2$type=="rm_imm_str0"), ]


ggplot(MHmelt3, aes(x=class, y=value, col=sampType))+geom_point()+facet_grid(~nam)+geom_errorbar(aes(ymin=lower, ymax=upper), width=.2)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+scale_color_manual(values=c("#E9A3C9", "#C51B7D", "#B5B2CC"))


#+scale_color_manual(values=c("#E9A3C9", "#C51B7D", "#B5B2CC"))



```

We can repeat the same analysis looking at immune-stroma mixing to see if this explains some of the spatial patterns that emerge. 


```{r}
MHmelt2=MHmelt[grep("str-imm", MHmelt$comp), ]
MHmelt2$nam=substr(MHmelt2$L1, 1, 1)
MHmelt2$class=substr(MHmelt2$L1, 2, 2)
MHmelt2$sampType=substr(MHmelt2$comp, 1, 1)
MHmelt2=MHmelt2[-which(MHmelt2$class=="I" & MHmelt2$sampType=="d"), ]
#MHmelt2$sampType2=paste(MHmelt2$comp, MHmelt$type)
MHmelt2=MHmelt2[which(MHmelt2$type=="keep all"), ]
MHmelt2$value[MHmelt2$L1=="1IDC.RData"]=NA

ggplot(MHmelt2, aes(x=class, y=value, col=sampType))+geom_point()+facet_grid(~nam)+geom_errorbar(aes(ymin=lower, ymax=upper), width=.2)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+scale_color_manual(values=c("#E9A3C9", "#C51B7D", "#B5B2CC"))


# plot stroma/immune mixing
plot(MHmelt3$value, MHmelt2$value, col=c(dcisidcCol, "purple")[factor(MHmelt2$sampType)], pch=19,
     xlab="immune-tumor mixing", ylab="immune-stroma mixing")

# add the two metrics together and determine the contribution of each to the model?
MHmodel1=data.frame(it=MHmelt3$value, is=MHmelt2$value, class=MHmelt3$sampType)
a1=glm(as.numeric(class)~it, data=MHmodel1[MHmodel1$class!="n", ])

ggplot(MHmelt3, aes(x=sampType, y=value, col=nam))+geom_boxplot(aes(col="black"))+
  geom_point(position = position_jitterdodge(jitter.width = 0, dodge.width =0.3), aes(color=nam))+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

ggplot(MHmelt2, aes(x=sampType, y=value, col=nam))+geom_boxplot(aes(col="black"))+
  geom_point(position = position_jitterdodge(jitter.width = 0, dodge.width =0.3), aes(color=nam))+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# do t.tests to compare different populations
# ignore the normal cases
wilcox.test(MHmelt3$value[which(MHmelt3$sampType!="n")]~MHmelt3$sampType[which(MHmelt3$sampType!="n")])
wilcox.test(MHmelt2$value[which(MHmelt2$sampType!="n")]~MHmelt2$sampType[which(MHmelt2$sampType!="n")])
# ignore dcis samples
wilcox.test(MHmelt3$value[which(MHmelt3$sampType!="d")]~MHmelt3$sampType[which(MHmelt3$sampType!="d")])
wilcox.test(MHmelt2$value[which(MHmelt2$sampType!="d")]~MHmelt2$sampType[which(MHmelt2$sampType!="d")])
# ignore idc samples
wilcox.test(MHmelt3$value[which(MHmelt3$sampType!="i")]~MHmelt3$sampType[which(MHmelt3$sampType!="i")])
wilcox.test(MHmelt2$value[which(MHmelt2$sampType!="i")]~MHmelt2$sampType[which(MHmelt2$sampType!="i")])

```


```{r}
#Also try stroma - tumor mixing: non-informative, similar to imm-tum mixing
MHmeltC=MHmelt[grep("str-tum", MHmelt$comp), ]
MHmeltC$nam=substr(MHmeltC$L1, 1, 1)
MHmeltC$class=substr(MHmeltC$L1, 2, 2)
MHmeltC$sampType=substr(MHmeltC$comp, 1, 1)
MHmeltC=MHmeltC[-which(MHmeltC$class=="I" & MHmeltC$sampType=="d"), ]
#MHmelt2$sampType2=paste(MHmelt2$comp, MHmelt$type)
MHmeltC=MHmeltC[which(MHmeltC$type=="rm_imm_str0"), ]
#MHmelt2$value[MHmelt2$L1=="1IDC.RData"]=NA

ggplot(MHmeltC, aes(x=class, y=value, col=sampType))+geom_point()+facet_grid(~nam)+geom_errorbar(aes(ymin=lower, ymax=upper), width=.2)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+scale_color_manual(values=c("#E9A3C9", "#C51B7D", "#B5B2CC"))


# plot stroma/immune mixing
plot(MHmelt3$value, MHmeltC$value, col=c("pink", "red", "purple")[factor(MHmelt2$sampType)], pch=19,
     xlab="immune-tumor mixing", ylab="immune-stroma mixing")
```