# Mutations

## All coding mutations 

Below is a plot of all coding mutations, color coded according to type of mutation. If a mutation is observed at exactly the same site, overlap in the plot is observed. 

In the recurrence cohort, very few mutations appear to be shared. Many 5DCIS mutations were also found in 5IDC for example, and a small proportion of mutations in cases 2, 8, and 9 were preserved. Case 1 and Case 3 had virtually no overlapping mutations.

```{r}
# shared DCIS mutations
AllGenes=lapply(Mutall2, function(x) (x[x$Variant_Classification%in%SeqOnt, c("Hugo_Symbol", "Variant_Classification","dbSNP_RS", "Protein_Change")]))

for (i in 1:12){
  AllGenes[[i]]$Variant_Classification=as.character(AllGenes[[i]]$Variant_Classification)
  AllGenes[[i]]$Variant_Classification <-ifelse(AllGenes[[i]]$dbSNP_RS!="", "rsSNP", AllGenes[[i]]$Variant_Classification)
}

AllGenesUn=unique(melt(AllGenes)$Hugo_Symbol)
Ta=melt(AllGenes)
Ta$ID=paste(Ta$Hugo_Symbol, Ta$Protein_Change)
lx1=which(duplicated(Ta[ ,c("ID", "L1")]))
if (length(lx1)>0){
Ta2=Ta[-which(duplicated(Ta[ ,c("ID", "L1")])), ]
} else {
  Ta2=Ta
}
N2=dcast(Ta2[ ,c("ID", "L1", "Variant_Classification")], ID~L1, value.var="Variant_Classification") #, value.var="Variant_Classification")
N2[which(N2==SeqOnt[3], arr.ind=T)]=1
N2[which(N2==SeqOnt[4], arr.ind=T)]=2
N2[which(N2==SeqOnt[1], arr.ind=T)]=3
N2[which(N2==SeqOnt[2], arr.ind=T)]=4
N2[which(N2==SeqOnt[5], arr.ind=T)]=5
N2[which(N2=="rsSNP", arr.ind=T)]=6

N2=N2[order(N2[ ,2], N2[,3], N2[,4], N2[,5],N2[ ,6], N2[,7], N2[,8], N2[,9], N2[ ,10], N2[,11], N2[ ,12],N2[ ,13], decreasing = T), ]
image(data.matrix(N2[ ,-1]), col=brewer.pal(6,"Spectral"), xaxt="none", yaxt="none")
axis(2, at=seq(0, 1, length=12), colnames(N2)[-1], las=2)
legend("bottomright", c("mis", "non", "fsd", "fsi","splice", "rsSNP"), col=brewer.pal(6,"Spectral"), pch=20)
```

Across all coding mutations: Most are missense (usually 50% upwards),

```{r, eval=F}
allOnto=unique(unlist(sapply(AllGenes, function(x) x$Variant_Classification)))
AllGSumm=sapply(AllGenes, function(x) summary(factor(x$Variant_Classification, levels=allOnto)))

AllGSumm2=t(AllGSumm)/(sum(AllGSumm))

AllGSumm2

barplot(t(AllGSumm2), beside=T, col=brewer.pal(6, "RdYlBu"))
legend("topleft", colnames(AllGSumm2), col=brewer.pal(6, "RdYlBu"), pch=20)
```


## Summary of oncogenic mutations

Below is an overlap of all coding mutations and CNVs found in common cancer-related genes.

Orange: CNV gain
Blue: CNV loss
Yellow: coding mutation
Green: a rsSNP site

Note that there are some mutations conserved in DCIS to IDC: for example, in Case 8 CDH1, TP53 are conserved. In Case 9, we see PIK3CA and GATA3 albeit at different sites. Case 5 also shares a coding TP53 mutation.


```{r}
## annotate Mutall2 if the gene is an oncogene

for (i in 1:length(Mutall2)){
  Mutall2[[i]]$Oncogene=ifelse(Mutall2[[i]]$Hugo_Symbol%in%OncoList$Gene.Symbol, 1, 0)
}

# refine the search to oncogenes
msearch=which(AllGenesUn%in%OncoList$Gene.Symbol)
msearch1=lapply(AllGenes, function(x) x[x$Hugo_Symbol%in%OncoList$Gene.Symbol, ])
msearchX=melt(msearch1)
colnames(msearchX)=c("genes", "VariantType",  "dbSNP_RS", "protein", "L1")
msearchX$mut=ifelse(msearchX$dbSNP_RS!="", paste("(", msearchX$protein, ")", msearchX$genes), as.character(msearchX$genes))
msearchX$value=ifelse(msearchX$VariantType=="rsSNP",  1, -1)
Dnew2=DnewOnc[which( DnewOnc$call!="0"), ] ## need to change this value!!
Dnew2=Dnew2[which(Dnew2$genes%in% c(as.character(msearchX$genes),breastGeneList )), ]
Dnew2$value=ifelse(Dnew2$call=="+", 2, -2)
Dnew2$L1=Dnew2$variable

msearchN=rbind(msearchX[ ,c("genes","L1","value")], Dnew2[ ,c("genes","L1","value" )])
lx1=which(duplicated(msearchN[ ,1:2]))
FNew3=acast(msearchN[-lx1, ], genes~L1)
Lx1=(abs(FNew3))
FNew3=FNew3[order(Lx1[ ,1], Lx1[ ,2], Lx1[ ,3], Lx1[ ,4], Lx1[ ,5],Lx1[ ,6],
                  Lx1[ ,7], Lx1[ ,8], Lx1[ ,9], Lx1[ ,10], Lx1[ ,11]),  ]

OldRN=rownames(FNew3)
lx1=msearchX$mut[match(rownames(FNew3), msearchX$genes)]

rownames(FNew3)[-which(is.na(lx1))]=na.omit(lx1)
MutCNVTabrecur=FNew3
image(FNew3, col=c("#9ecae1", "#fed976","white", "#addd8e", "#fb6a4a"), xaxt="n", yaxt="n")

axis(1, at=seq(0, 1, length=nrow(FNew3)), rownames(FNew3), las=2, cex=0.6,  cex.axis=0.7)
axis(2, at=seq(0, 1, length=12), paste(rep(c(1:3, 5, 8:9), each=2), c("DCIS", "IDC")), las=2)

```

The expression profile corresponding to the above plot. Note that expression values are z-scaled, so ERBB2 is high in case 9 relative to the other cases, as is GATA3. PTEN which is mutated in Case 2 has lower expression compared to the other samples?

```{r}
MTable=normalizedTableVSD[(match(OldRN, rownames(normalizedTableVSD))), na.omit(match(colnames(FNew3), colnames(normalizedTableVSD)))]
MTable2=scale(t(MTable))
heatmap.2(MTable2[-(grep("str", colnames(normalizedTableVSD))) ,], col=RdBu[11:1], trace="none", Colv = NA, Rowv = NA)
```

## Abba and Lesurf sets

Below is a similar plot for a combined Lesurf and Abba cohort, highlighting genes which are gained and lost and also have mutations. Erbb2, Myc are commonly gained for example.

Orange: CNV gain
Blue: CNV loss
Yellow: coding mutation
Green: a rsSNP site

```{r, fig.height=8}
AAllGenes=lapply(AMutall, function(x) (x[x$Variant_Classification%in%SeqOnt, c("Hugo_Symbol", "Variant_Classification","dbSNP_RS", "Protein_Change")]))

for (i in 1:length(AMutall)){
  AAllGenes[[i]]$Variant_Classification=as.character(AAllGenes[[i]]$Variant_Classification)
  AAllGenes[[i]]$dbSNP_RS=ifelse(is.na(AAllGenes[[i]]$dbSNP_RS)==T, "", as.character(AAllGenes[[i]]$dbSNP_RS))
  AAllGenes[[i]]$Variant_Classification <-ifelse(AAllGenes[[i]]$dbSNP_RS!="", "rsSNP", AAllGenes[[i]]$Variant_Classification)
}

AAllGenesUn=unique(melt(AAllGenes)$Hugo_Symbol)
Ta=melt(AAllGenes)
Ta2=Ta[-which(duplicated(Ta[ ,c("Hugo_Symbol", "L1")])), ]
Ta2$Hugo_Symbol=factor(Ta2$Hugo_Symbol)

N2=dcast(Ta2, Hugo_Symbol~L1, value.var="Variant_Classification") #, value.var="Variant_Classification")
N2[which(N2=="Missense_Mutation", arr.ind=T)]=1
N2[which(N2=="Nonsense_Mutation", arr.ind=T)]=2
N2[which(N2=="Frame_Shift_Del", arr.ind=T)]=3
N2[which(N2=="Frame_Shift_Ins", arr.ind=T)]=4
N2[which(N2=="Splice_Site", arr.ind=T)]=5
N2[which(N2=="rsSNP", arr.ind=T)]=6
rownames(N2)=N2[ ,1]

N2[which(is.na(N2)==T, arr.ind=T)]=0


for (i in 1:length(AMutall)){
  AMutall[[i]]$Oncogene=ifelse(AMutall[[i]]$Hugo_Symbol%in%OncoList$Gene.Symbol, 1, 0)
}

# refine the search to oncogenes
msearch=which(AAllGenesUn%in%OncoList$Gene.Symbol)
msearch1=lapply(AAllGenes, function(x) x[x$Hugo_Symbol%in%OncoList$Gene.Symbol, ])
msearchX=melt(msearch1)
colnames(msearchX)=c("Gene", "VariantType",  "dbSNP_RS", "protein", "variable")
msearchX$mut=ifelse(msearchX$dbSNP_RS!="", paste("(", msearchX$protein, ")", msearchX$Gene), as.character(msearchX$Gene))
msearchX$call=ifelse(msearchX$VariantType=="rsSNP", -1, -2)
## extract the amp/loss information
# may need to change threshold here
ADnew2=AbbaDnewOnc[which(AbbaDnewOnc$call!="0"), ] ## need to change this value!!
ADnew2=ADnew2[which(ADnew2$Gene%in% c(as.character(msearchX$Gene),breastGeneList )), ]
ADnew2$call=ifelse(ADnew2$call=="+", 2, -3)
ADnew2$L1=ADnew2$pat

## also include the Lesurf CNV set
LBreastmeltB=LBreastmelt[which(LBreastmelt$Gene%in% c(as.character(msearchX$Gene),breastGeneList)),]

msearchN=rbind(msearchX[ ,c("Gene","variable", "call")], ADnew2[ ,c("Gene","variable","call" )], LBreastmeltB[ ,c("Gene", "variable", "call")])
lx1=which(duplicated(msearchN[ ,1:2]))
FNew3=acast(msearchN[-lx1, ], Gene~variable)

FNew4=melt(FNew3)
FNew4$subtype=AbbaClin$`PAM50 call`[match(FNew4$Var2, AbbaClin$SampleID)]

lidx=c(grep("EuCan", FNew4$Var2), grep("FW",  FNew4$Var2))

FNew4$subtype[lidx]=dcis_annot$PAM50[match(FNew4$Var2[lidx], rownames(dcis_annot))]
FNew4$chrom=as.numeric(as.character(seqnames(tx37)[match(FNew4$Var1, tx37$HUGO)]))
FNew4$Loc=start(ranges(tx37))[match(FNew4$Var1, tx37$HUGO)]

FNew4=FNew4[order(FNew4$Loc), ]
FNew4$Var1=factor(FNew4$Var1, levels=unique(FNew4$Var1))

p=ggplot(FNew4, aes(x=Var1, y=Var2, fill=factor(value)))+geom_tile()+facet_grid(subtype~chrom,scales="free", space="free_x")+
  scale_fill_manual(values = c("#9ecae1", "#fed976", "#addd8e",  "#fb6a4a"), na.value="white")+theme(axis.text.x=element_text(angle=90, vjust=.8, hjust=1.01))+ggtitle(label="Abba set")

gp <- ggplotGrob(p)
# get gtable columns corresponding to the facets (5 & 9, in this case)
facet.rows <- gp$layout$t[grepl("panel", gp$layout$name)]

# get the number of unique x-axis values per facet (1 & 3, in this case)
x.var <- sapply(ggplot_build(p)$layout$panel_scales_y,
                function(l) length(l$range$range))

# change the relative widths of the facet columns based on
# how many unique x-axis values are in each facet
gp$heights[facet.rows] <- gp$heights[facet.rows] * x.var

# plot result
grid::grid.draw(gp)


```

Below we calculate exactly the number of cases which have an amplification ofor loss for key oncogenes of interest:

<span style="color:red"> calculate proportion for the Lesurf cohort too </span>

For HER2, `r length(which(ADnew2[(ADnew2$Gene=="ERBB2"), "call"]=="2"))` samples in the Abba cohort have gains.
`r length(which(LBreastmeltB[LBreastmeltB$Gene=="ERBB2", "call"]=="2"))` samples in the Lesurf cohort have gains.

PIK3CA: `r length(which(msearchX$call[msearchX$Gene=="PIK3CA"]==(-1)))` samples in the Abba coort have mutations. Note that one sample (T14) has two different PIK3CA mutations.

TP53: `r length(which(msearchX$call[msearchX$Gene=="TP53"]<0))` samples in the Abba cohort have mutations

GATA3: `r length(which(msearchX$call[msearchX$Gene=="GATA3"]<0))` samples in the Abba cohort have mutations

```{r}
print('Abba cases HER2')
table(ADnew2[(ADnew2$Gene=="ERBB2"), "call"])
print('Abba cases PIk3CA')
table(msearchX$call[msearchX$Gene=="PIK3CA"], msearchX$variable[msearchX$Gene=="PIK3CA"])
print('Abba cases TP53')
table(msearchX$call[msearchX$Gene=="TP53"], msearchX$variable[msearchX$Gene=="TP53"])
print('Abba cases GATA3')
table(msearchX$call[msearchX$Gene=="GATA3"], msearchX$variable[msearchX$Gene=="GATA3"])


print('Lesurf set ERBB2 calls')
table(LBreastmeltB[which(LBreastmeltB$Gene=="ERBB2"), "call"])
```
